---
title: "Landing distance of a commercial flight"
author: "by Hasnat Tahir"
output: rmarkdown::github_document

---
```{r}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
```


Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
library(ggplot2)
library(DataExplorer)
library(broom)
library(dplyr)
library(MASS)
```


Step-1
```{r}
faa1 = read_excel('D:\\University\\Pet projects\\Airline\\FAA1.xls')

faa2 = read_excel('D:\\University\\Pet projects\\Airline\\FAA2.xls')

dim(faa1)
dim(faa2)

```
Both FAA1 and FAA2 data sets have been imported.

Step-2
```{r}
str(faa1)
str(faa2)
```
In FAA1 data set there are 800 observations and 8 variables.
In FAA2 data set there are 150 observations and 7 variables.
The only difference is that 'duration' variable in not present in FAA2 data set.

Step-3
```{r}
faa2$duration = NA # adds a duration column to facilitate merging

faa = rbind(faa1, faa2) #merging the two data sets

dim(faa)

#duplicate values
faa_uniq = faa %>% distinct(aircraft,distance,no_pasg,speed_ground,speed_air,height,pitch,.keep_all=TRUE)

dim(faa_uniq)
```
100 duplicate values were observed. Yes, I got rid of those duplicate values as they are a result of some error and they might bias the prediction model.

Step-4
```{r}
str(faa_uniq)

summary(faa_uniq)
```
The combined data set 'faa' has 950 observations and 8 variables.

### Step-5
Key Observations:

* The FAA2 data set doesn't have the duration variable.
* The Height variable has some negative values.
* The speed_air variable has approx 75% of the observation values missing.
* Based on the problem statement, speed_ground and speed_air some values  <30 mph and >140 mph that are considered abnormal.
* Duration variable too has some abnormal values that needs to be taken care before doing any analysis.

Step-6
```{r}
#Negative Height and abnormal height
Abnormal_Height = faa_uniq[faa_uniq$height < 6,]
Abnormal_Height # 10 values


#Abnormal Speeds
Abnormal_speed = faa_uniq[(faa_uniq$speed_ground >140 | faa_uniq$speed_ground < 30),]
Abnormal_speed # 3 abnormal values

Abnormal_speed2 = faa_uniq[(faa_uniq$speed_air > 140 | faa_uniq$speed_air < 30) & (is.na(faa_uniq$speed_air) ==FALSE),]
Abnormal_speed2 # 1 values

#Duration
Abnormal_duration = faa_uniq[(faa_uniq$duration< 40) & (is.na(faa_uniq$duration) ==FALSE),]
Abnormal_duration # 5 values

```
* Height has 10 abnormal values.
* speed_ground has 3 abnormal values.
* speed_air has 1 abnormal values.
* duration has 5 abnormal values.

```{r}
#removing abnormal values

clean_faa = filter(faa_uniq, (faa_uniq$duration > 40 | is.na(faa_uniq$duration == T)))
clean_faa1 = filter(clean_faa, ((clean_faa$speed_air >=30 & clean_faa$speed_air<=140) | is.na(clean_faa$speed_air == T)))

clean_faa2 = filter(clean_faa1, ((clean_faa1$speed_ground >=30 & clean_faa1$speed_ground<=140)))

clean_faa3 = filter(clean_faa2, clean_faa2$height >= 6)
                              
dim(clean_faa3)

```
All the abnormal values have been removed.


Step-7
```{r}
str(clean_faa3)
summary(clean_faa3)
```


Step-8
```{r}
plot_histogram(clean_faa3)
plot_bar(clean_faa3$aircraft)
```
Above are the histograms for all the variables.

### Step-9

* speed_air and duration variables still have missing values.
* height, speed_air, speed_ground and duration variables had abnormal values that have been removed.
* The final clean data set has 832 observations and 8 variables.

# Part 2

### Step-1
```{r}
#binary variables
clean_faa3$long.landing = ifelse(clean_faa3$distance > 2500, 1, 0)
clean_faa3$risky.landing = ifelse(clean_faa3$distance > 3000, 1, 0)

#deleting distance variable
colnames(clean_faa3)

new_faa = clean_faa3[,-8]
head(new_faa)

#aircraft to binary
new_faa$aircraft = ifelse(new_faa$aircraft == "airbus", 1,0) 

summary(new_faa$aircraft)
typeof(new_faa$aircraft)
```
The binary variables long.landing and risky.landing have been created.

### Step-2
```{r}
#pie chart showing distribution of long landing
piec = table(new_faa$long.landing)
lbl = c("Normal landing", "Long landing")

pct = c(round(100-round(sum(new_faa$long.landing==1)/length(new_faa))), round(sum(new_faa$long.landing==1)/length(new_faa)))
pct
lbl = paste(lbl, pct) 
lbl = paste(lbl,"%",sep="")
pie(piec, labels = lbl, main = "Long Landing")

```
The above pie charts shows the distribution of Long landing and risky landing.

### Step-3:
```{r}
#Creating table 2
reg_glm = function(dat,vari){
  z = glm(dat$long.landing~vari, family = binomial, data = dat)
  x = tidy(z)$p.value[2]
  y = tidy(z)$estimate[2]
  return(c(y,x))
}

r1 = reg_glm(new_faa, new_faa$aircraft)
r2 = reg_glm(new_faa, new_faa$duration)
r3 = reg_glm(new_faa, new_faa$no_pasg)
r4 = reg_glm(new_faa, new_faa$speed_ground)
r5 = reg_glm(new_faa, new_faa$speed_air)
r6 = reg_glm(new_faa, new_faa$height)
r7 = reg_glm(new_faa, new_faa$pitch)

tabl = rbind(r1,r2,r3,r4,r5,r6,r7)


colnames(tabl) = c('Coefficients', 'p-value')

tabl = as.data.frame(tabl)
tabl['Direction'] = ifelse(tabl$Coefficients < 0, "Negative", "Positive")

col_nam = c("aircraft", "duration","no_pasg","speed_ground","speed_air","height","pitch")

tabx = cbind(col_nam,tabl)

tabx['Odds ratio'] = exp(tabx$Coefficients)

table = tabx[order(-tabx$`Odds ratio`, tabx$`p-value`),]
table

```
The above table has been ordered by descending odds ratio and ascending p-values.

Observations:

* Coefficients for aircraft, duration, and no_pasg are negative
* speed_air has the highest odds ratio 
* variables height, duration, and no_pasg are insignificant based on thier p-values.  

### Step-4:
```{r jitter long}
#jitter plots
jp = function(vari,vname){
  plot(jitter(new_faa$long.landing,0.1)~jitter(vari), new_faa,xlab = vname, ylab = "long landing", pch = ".")
}

jp(new_faa$speed_ground, "Speed ground")
jp(new_faa$speed_air, "Speed air")
jp(new_faa$height, "Height")
jp(new_faa$duration, "Duration")
jp(new_faa$no_pasg, "No. of passengers")
jp(new_faa$pitch, "Pitch")
jp(new_faa$aircraft, "Aircraft")
```
Observations:

* As speed_ground or speed_air crosses a certain value the the number of long landing cases increases.
* Based on the analysis in step-3&4 and step 16 of part 1, following variables have been selected for further analysis:
  + Speed_air
  + aircraft
  + Height
  + pitch

### Step 5: Fitting model with the selected variables:
```{r}
#fitting model
full_model = glm(long.landing~speed_air+aircraft+height+pitch, family = binomial, data= new_faa)
summary(full_model)
```
* The above model gives an AIC of 44.856.
* Coefficient of pitch variable comes out to be insignificant in the model
* Aircraft variable has the biggest coefficient value


```{r}
#removing NA values
new_faac = na.omit(new_faa)
dim(new_faa)
dim(new_faac)

```
Null values have been removed to perform Stepwise variable selection.  

### Step-6:
Forward stepwise selection:
```{r Forward Stepwise , warning=FALSE}
#forward selection using AIC
null_mod = glm(long.landing~1, family = "binomial", data = new_faac)
full_mod = glm(long.landing~., family = "binomial", data = new_faac[,-9])

mod_step_f = step(null_mod,
                  scope = list(lower= null_mod, upper = full_mod),
                  direction = "forward")
summary(mod_step_f)
```
As per the forward stepwise selection method using AIC following variables has been selected based on AIC:

* speed_air
* aircraft
* height
* pitch

Yes, the results is consistent with the variable selection done after step 3. 

### Step-7:
```{r Forward Stepwise (BIC), warning=FALSE}
#forward selection using BIC
null_mod0 = glm(long.landing~1, family = "binomial", data = new_faac)
full_mod0 = glm(long.landing~., family = "binomial", data = new_faac[,-9])

mod_step_f0 = step(null_mod,
                  scope = list(lower= null_mod0, upper = full_mod0),
                  direction = "forward", k = log(length(new_faac$long.landing)))
summary(mod_step_f0)

table.selected = tidy(mod_step_f0)
```
As per the forward stepwise selection method using BIC following variables has been selected based on AIC:

* speed_air
* aircraft
* height

The only difference between variable selection using AIC and BIC is pitch was not included in the later method.

### Step 8:

**Model** : The ouput model of Forward stepwise using BIC has been selected as it gives the simplest model with almost same AIC value. *(long.landing ~ speed_air + aircraft + height)* 

**Table**: 'table.selected' containing summary of selected model. So, that more about the selected model can be presented. 

**Figures**: AIC and Residual Deviance of the selected model.

* The risk factors associated with long landings are speed_air, aircraft and height.
* If speed air is above a certain value the chances of long landing increses substantially.
* As per the modeling, aircraft too has some influence on long landing, but it needs to be studied further.



### Step 9:
```{r}
#pie chart showing distribution of risky landing
piec1 = table(new_faa$risky.landing)
lbl1 = c("Normal landing", "Risky landing")
pct1 = c(100-round(sum(new_faa$risky.landing==1)/length(new_faa)), round(sum(new_faa$risky.landing==1)/length(new_faa)))
pct1
lbl1 = paste(lbl1, pct) 
lbl1 = paste(lbl1,"%",sep="")
pie(piec1, labels = lbl1, main = "Risky Landing" )

```

Of all the observations 12 percent have been categorised as risky landing.

```{r table risky}
#Creating table 2
reg_glm1 = function(dat,vari){
  z = glm(risky.landing~vari, family = binomial, data = dat)
  x = tidy(z)$p.value[2]
  y = tidy(z)$estimate[2]
  return(c(y,x))
}

g1 = reg_glm1(new_faa, new_faa$aircraft)
g2 = reg_glm1(new_faa, new_faa$duration)
g3 = reg_glm1(new_faa, new_faa$no_pasg)
g4 = reg_glm1(new_faa, new_faa$speed_ground)
g5 = reg_glm1(new_faa, new_faa$speed_air)
g6 = reg_glm1(new_faa, new_faa$height)
g7 = reg_glm1(new_faa, new_faa$pitch)

tabl1 = rbind(g1,g2,g3,g4,g5,g6,g7)


colnames(tabl1) = c('Coefficients', 'p-value')

tabl1 = as.data.frame(tabl1)
tabl1['Direction'] = ifelse(tabl1$Coefficients < 0, "Negative", "Positive")

col_nam = c("aircraft", "duration","no_pasg","speed_ground","speed_air","height","pitch")

tabx1 = cbind(col_nam,tabl1)

tabx1['Odds ratio'] = exp(tabx1$Coefficients)

table1 = tabx1[order(-tabx1$`Odds ratio`, tabx1$`p-value`),]
table1
```
The above table has been ordered by descending odds ratio and ascending p-values.

Observations:

* Coefficients for aircraft, duration, and no_pasg are negative
* speed_air has the highest odds ratio 
* variables pitch, height, duration, and no_pasg are insignificant based on thier p-values. 

```{r}
#jitter plots
jp1 = function(vari,vname){
  plot(jitter(new_faa$risky.landing,0.1)~jitter(vari), new_faa,xlab = vname, ylab = "risky landing", pch = ".")
}

jp1(new_faa$speed_ground, "Speed ground")
jp1(new_faa$speed_air, "Speed air")
jp1(new_faa$height, "Height")
jp1(new_faa$duration, "Duration")
jp1(new_faa$no_pasg, "No. of passengers")
jp1(new_faa$pitch, "Pitch")
jp1(new_faa$aircraft, "Aircraft")
```

Observations:

* As speed_ground or speed_air crosses a certain value the the number of long landing cases increases.

\ 
```{r}
full_model1 = glm(risky.landing~speed_air+aircraft, family = binomial, data= new_faa)
summary(full_model1)
```

```{r}
#AIC
null_mod1 = glm(risky.landing~1, family = "binomial", data = new_faac)
full_mod1 = glm(risky.landing~., family = "binomial", data = new_faac[,-8])

mod_step_f1 = step(null_mod1,
                  scope = list(lower= null_mod1, upper = full_mod1),
                  direction = "forward")
summary(mod_step_f1)
```
As per the forward stepwise selection method using AIC following variables has been selected based on AIC:

* speed_air
* aircraft

```{r}
#forward selection using BIC
null_mod2 = glm(risky.landing~1, family = "binomial", data = new_faac)
full_mod2 = glm(risky.landing~., family = "binomial", data = new_faac[,-8])

mod_step_f2 = step(null_mod2,
                  scope = list(lower= null_mod2, upper = full_mod2),
                  direction = "forward", k = log(length(new_faac$risky.landing)))
summary(mod_step_f2)

table.risky = tidy(mod_step_f2)
```
As per the forward stepwise selection method using BIC following variables has been selected based on BIC:

* speed_air
* aircraft

The variables selected are similar to what was selected using AIC.

### Step 10

**Model** : The ouput model of Forward stepwise using BIC has been selected. *(risky.landing ~ speed_air + aircraft)* 

**Table**: 'table.risky' containing summary of selected model. So, that more about the selected model can be presented. 

**Figures**: AIC and Residual Deviance of the selected model.

* The risk factors associated with risky landings are speed_air and aircraft.
* If speed air is above a certain value the chances of risky landing increses substantially.
* As per the modeling, aircraft too has some influence on risky landing, but it needs to be studied further.


### Step 11
Difference between selected models for long landing and risky landing

* Risky landing model has all the same variables as long landing model except height.
* Based on the model, aircraft variable is more significant in long landing model.
* BIC value for risky landing model is better than that of long landing model.

### Step 12
```{r}
glm.long = glm(long.landing~speed_air+aircraft+height, family = binomial, data = new_faac)
summary(glm.long)

pred.prob1 = predict(glm.long, type = "response")

#pred.out1 = ifelse(pred.prob1 < 0.5,"no","yes") 
new_faac1 = data.frame(new_faac,pred.prob1)


thresh = seq(0.01, 0.5, 0.01)
sensitivity1 = specificity1 = rep(NA,length(thresh))
for(j in seq(along= thresh)){
pp1 = ifelse(new_faac1$pred.prob1<thresh[j],"no","yes")
xx1 = xtabs(~long.landing+pp1,new_faac1)
specificity1[j] = xx1[1,1]/(xx1[1,1]+xx1[1,2])
sensitivity1[j] = xx1[2,2]/(xx1[2,1]+xx1[2,2])
}
#par(mfrow=c(1,2))
#matplot(thresh,cbind(sensitivity,specificity),type="l",xlab="Threshold",ylab="Proportion",lty=1:2)
#plot(1-specificity1,sensitivity1,type="l");abline(0,1,lty=2)
```

```{r}
glm.risky = glm(risky.landing~speed_air+aircraft, family = binomial(link = "logit"), data = new_faac)
summary(glm.risky)

pred.prob2 = predict(glm.risky, type = "response")


new_faac2 = data.frame(new_faac,pred.prob2)


thresh = seq(0.01, 0.5, 0.01)
sensitivity = specificity = rep(NA,length(thresh))
for(j in seq(along= thresh)){
pp = ifelse(new_faac2$pred.prob2<thresh[j],"no","yes")
xx = xtabs(~risky.landing+pp,new_faac2)
specificity[j] = xx[1,1]/(xx[1,1]+xx[1,2])
sensitivity[j] = xx[2,2]/(xx[2,1]+xx[2,2])
}


#both ROC curves 
plot(1-specificity1,sensitivity1,type="l", col = "Red", xlab="1-specificity", ylab = "Sensitivity", main = "ROC curves for Long.landing(Red) & Risky.landing(Black)") #long landing
lines(1-specificity,sensitivity) #risky landing
```

The above plot shows ROC curve for the slected long and risky landing models

There is a difference, area under the ROC curve of long.landing is more as compared to ROC curve for risky landing. 

### Step 13
```{r prediction}
#long
new_pred = data.frame(aircraft=0, duration = 200, no_pasg=80, speed_ground=115, speed_air=120, height = 40, pitch= 4)
new_pred

#predict 
predict_1 = predict(mod_step_f0, newdata = new_pred, type = "response", se=T)

#confidence interval

c(predict_1$fit-1.96*predict_1$se.fit,predict_1$fit+1.96*predict_1$se.fit)

#risky
#predict 
predict_1 = predict(mod_step_f2, newdata = new_pred, type = "response", se=T)

#confidence interval

c(predict_1$fit-1.96*predict_1$se.fit,predict_1$fit+1.96*predict_1$se.fit)

```
There is a very high probability of being a long as well as risky landing for the given commercial plane.

### Step 14
```{r}
#probit model

risky.probit = glm(risky.landing~speed_air+aircraft, family = "binomial"(link = "probit"), data = new_faac)
#summary(risky.probit)

#hazard model
risky.hazard = glm(risky.landing~speed_air+aircraft, family = "binomial"(link = "cloglog"), data = new_faac)
#summary(risky.hazard)

# logit model
#summary(glm.risky)

```

compare above probit and harzard model with logit model
```{r}
round(coef(glm.risky),3)
round(coef(risky.probit),3)
round(coef(risky.hazard),3)
```
The above output shows that coefficients of speed air and aircraft are bigger as compared to the probit and hazard model. 

### Step 15
```{r}
#probit
pred.prob3 = predict(risky.probit, type = "response")


new_faac3 = data.frame(new_faac,pred.prob3)


thresh = seq(0.01, 0.5, 0.01)
sensitivity2 = specificity2 = rep(NA,length(thresh))
for(j in seq(along= thresh)){
pp2 = ifelse(new_faac3$pred.prob3<thresh[j],"no","yes")
xx2 = xtabs(~risky.landing+pp2,new_faac3)
specificity2[j] = xx2[1,1]/(xx2[1,1]+xx2[1,2])
sensitivity2[j] = xx2[2,2]/(xx2[2,1]+xx2[2,2])
}

#hazard
pred.prob4 = predict(risky.hazard, type = "response")


new_faac4 = data.frame(new_faac,pred.prob4)


thresh = seq(0.01, 0.5, 0.01)
sensitivity3 = specificity3 = rep(NA,length(thresh))
for(j in seq(along= thresh)){
pp3 = ifelse(new_faac4$pred.prob4<thresh[j],"no","yes")
xx3 = xtabs(~risky.landing+pp3,new_faac4)
specificity3[j] = xx3[1,1]/(xx3[1,1]+xx3[1,2])
sensitivity3[j] = xx3[2,2]/(xx3[2,1]+xx3[2,2])
}


#both ROC curves 
plot(1-specificity,sensitivity,type="l", col = "Red", xlab="1-specificity", ylab = "Sensitivity", main = "ROC curves for risky.landing(logit = Red, probit = black, hazard = green") #logit
lines(1-specificity2,sensitivity2) #probit
lines(1-specificity3,sensitivity3, col = "green") #hazard
```

It is clear from the above ROC curve that Probit model has the highest area under the curve. 

### Step 16
```{r}
#logit
predval = sapply(list(glm.risky), fitted)
colnames(predval) =  "logit"

predv = as.data.frame(predval)

order(-predv) #index

top.5.logit = new_faac2[c(49,116,20,58,126), c(1,10)] # 0 boeing and 1 is airbus 
top.5.logit

#probit
predval1 = sapply(list(risky.probit), fitted)
colnames(predval1) =  "Probit"

predv1 = as.data.frame(predval1)

order(-predv1) #index

top.5.probit = new_faac3[c(6,10,11,17,20), c(1,10)] # 0 boeing and 1 is airbus
top.5.probit
#hazard
predval2 = sapply(list(risky.hazard), fitted)
colnames(predval2) =  "Hazard"

predv2 = as.data.frame(predval2)

order(-predv2) #index

top.5.hazard = new_faac4[c(6,10,11,17,20), c(1,10)] # 0 boeing and 1 is airbus
top.5.hazard
```
Based of the above analayses:

* For logit model out of top 5 risky landing 4 flights are Boeing and 1 is Airbus.
* For Probit and Hazard model all top 5 risky landings are for Boeing.
* All the observations have a predicted probability of 1.

### Step 17
```{r}
#predict probit 
predict.probit = predict(risky.probit, newdata = new_pred, type = "response", se=T)
predict.probit
#confidence interval

c(predict.probit$fit-1.96*predict.probit$se.fit,predict.probit$fit+1.96*predict.probit$se.fit)

#predict clog log 
predict.hazard = predict(risky.hazard, newdata = new_pred, type = "response", se=T)
predict.hazard
#confidence interval

c(predict.hazard$fit-1.96*predict.hazard$se.fit,predict.hazard$fit+1.96*predict.hazard$se.fit)
```
* The prediction for the flights described using Probit and Hazard model are same and is equal to 1.
* The prediction obtained is same as what we got in step 13, however, the confidence interval in case of logit model is (0.9999997 - 1.0000002), whereas for other to, models its (1-1).
